# R and All that {-}



## Linear Algebra {-}

### Vectors {-}

Define the vectors $u = \begin{pmatrix} 1 \\2 \\3 \end{pmatrix}$, $v = \begin{pmatrix} 4\\5\\6 \end{pmatrix}$, and the scalar $c = 2$. Calculate the following:

1. $u + v$
2. $cv$
3. $u \cdot v$

If you are having trouble with these problems, please review Section \@ref(vector-def) "Working with Vectors" in Chapter \@ref(linearalgebra). 


Are the following sets of vectors linearly independent?

1. $u = \begin{pmatrix} 1\\ 2\end{pmatrix}$, $v = \begin{pmatrix} 2\\4\end{pmatrix}$


2. $u = \begin{pmatrix} 1\\ 2\\ 5 \end{pmatrix}$, $v = \begin{pmatrix} 3\\ 7\\ 9 \end{pmatrix}$


3. $a = \begin{pmatrix} 2\\ -1\\ 1 \end{pmatrix}$, $b = \begin{pmatrix} 3\\ -4\\ -2 \end{pmatrix}$, $c = \begin{pmatrix} 5\\ -10\\ -8 \end{pmatrix}$ (this requires some guesswork)


If you are having trouble with these problems, please review Section \@ref(linearindependence). 


### Matrices {-}

\[{\bf A}=\begin{pmatrix}
			7 & 5 & 1 \\
			11 & 9 & 3 \\ 
			2 & 14 & 21 \\ 
			4 & 1 & 5
		\end{pmatrix}\]
		
		
What is the dimensionality of matrix ${\bf A}$?

What is the element $a_{23}$ of ${\bf A}$?

Given that 
		
\[{\bf B}=\begin{pmatrix}
			1 & 2 & 8 \\
			3 & 9 & 11 \\ 
			4 & 7 & 5 \\ 
			5 & 1 & 9
		\end{pmatrix}\]
		
What is ${\bf A}$ + ${\bf B}$?


Given that 
		
\[{\bf C}=\begin{pmatrix}
			1 & 2 & 8 \\
			3 & 9 & 11 \\ 
			4 & 7 & 5 \\ 
		\end{pmatrix}\]
		
What is ${\bf A}$ + ${\bf C}$?



Given that 

\[c = 2\]

What is $c$${\bf A}$?

If you are having trouble with these problems, please review Section \@ref(matrixbasics). 



## Operations {-}

### Summation {-}

Simplify the following

1. \(\sum\limits_{i = 1}^3 i\)

2. \(\sum\limits_{k = 1}^3(3k + 2)\)

3. \(\sum\limits_{i= 1}^4 (3k + i + 2)\)


### Products {-}

1. \(\prod\limits_{i= 1}^3 i\)

2. \(\prod\limits_{k=1}^3(3k + 2)\)


To review this material, please see Section \@ref(sum-notation).


### Logs and exponents {-}

Simplify the following

1.  \(4^2\)
2.  \(4^2 2^3\)
3.  \(\log_{10}100\)
4.  \(\log_{2}4\)
5.  \(\log e\), where $\log$ is the natural log (also written as $\ln$) -- a log with base $e$, and $e$ is Euler's constant
6.  \(e^a e^b e^c\), where $a, b, c$ are each constants
7.  \(\log 0\)
8.  \(e^0\)
9.  \(e^1\)
10. \(\log e^2\)

To review this material, please see Section \@ref(logexponents)



## Limits {-}

Find the limit of the following.

1. $\lim\limits_{x \to 2} (x - 1)$
2. $\lim\limits_{x \to 2} \frac{(x - 2) (x - 1)}{(x - 2)}$
3. $\lim\limits_{x \to 2}\frac{x^2 - 3x + 2}{x- 2}$

To review this material please see Section \@ref(limitsfun)


## Calculus {-}

For each of the following functions $f(x)$, find the derivative $f'(x)$ or $\frac{d}{dx}f(x)$

1. $f(x)=c$
2. $f(x)=x$
3. $f(x)=x^2$
4. $f(x)=x^3$
5. $f(x)=3x^2+2x^{1/3}$
6. $f(x)=(x^3)(2x^4)$

For a review, please see Section \@ref(derivintro) - \@ref(derivpoly)

## Optimization {-}

For each of the followng functions $f(x)$, does a maximum and minimum exist in the domain $x \in \mathbf{R}$? If so, for what are those values and for which values of $x$?

1. $f(x) = x$
2. $f(x) = x^2$
3. $f(x) = -(x  - 2)^2$

If you are stuck, please try sketching out a picture of each of the functions.
	

## Probability {-}

1. If there are 12 cards, numbered 1 to 12, and 4 cards are chosen, how many distinct possible choices are there? (unordered, without replacement)
2. Let $A = \{1,3,5,7,8\}$ and $B = \{2,4,7,8,12,13\}$. What is $A \cup B$? What is $A \cap B$? If $A$ is a subset of the Sample Space $S = \{1,2,3,4,5,6,7,8,9,10\}$, what is the complement $A^C$? 
3. If we roll two fair dice, what is the probability that their sum would be 11?
4. If we roll two fair dice, what is the probability that their sum would be 12?


For a review, please see Sections \@ref(setoper) - \@ref(probdef). 


http://www.ynufe.edu.cn/pub/tsxy/jhzhao/teach/CS/compt_stat.htm  



```{r}


## ----Getting Started with R----------------------------------------------
1/sqrt(2*pi) * exp(-2)
dnorm(2)

## ------------------------------------------------------------------------
seq(0, 3, 0.5)
x <- seq(0, 3, 0.5)

 

## ----eval=FALSE----------------------------------------------------------
## help ( density )
## # copy and paste the lines below from the help page
## # The Old Faithful geyser data
## d <- density ( faithful $ eruptions , bw = "sj")
## d
## plot (d)

## ----Functions-----------------------------------------------------------
sumdice <- function (n) {
  k <- sample (1:6 , size =n, replace = TRUE )
  return (sum (k))
}

## ------------------------------------------------------------------------
sumdice (2) #to print the result at the console
a <- sumdice (100) #to store the result rather than print it
a / 100 #we expect the mean for 100 dice to be close to 3.5

## ------------------------------------------------------------------------
sumdice <- function (n)
  sum ( sample (1:6 , size =n, replace = TRUE ))

## ------------------------------------------------------------------------
sumdice <- function (n, sides = 6) {
  if ( sides < 1) return (0)
  k <- sample (1: sides , size =n, replace = TRUE )
  return (sum (k))
}
sumdice (5) # default 6 sides
sumdice (n=5, sides =4) #4 sides

## ----Data Frame----------------------------------------------------------
names ( iris )
table ( iris $ Species )
w <- iris [[2]] # Sepal . Width
mean (w)

## ------------------------------------------------------------------------
attach(iris)
summary ( Petal.Length [51:100]) # versicolor petal length
with(iris, summary(Petal.Length[51:100]))

## ------------------------------------------------------------------------
by(iris[,1:4],Species,colMeans)

## ----Arrays--------------------------------------------------------------
x <- 1:24 # vector
dim(x) <- length(x) # 1 dimensional array
matrix(1:24, nrow=4, ncol=6) # 4 by 6 matrix
x <- array(1:24, c(3, 4, 2))

## ----Example 1.3, A: 2 × 2 zero matrix-----------------------------------
A <- matrix (0, nrow =2, ncol =2)
A <- matrix (c(0, 0, 0, 0), nrow =2, ncol =2)
A <- matrix (0, 2, 2)

## ------------------------------------------------------------------------
A <- matrix (1:8 , nrow =2, ncol =4)

## ------------------------------------------------------------------------
x <- as.matrix ( iris [ ,1:4]) # all rows of columns 1 to 4
mean (x[ ,2]) # mean of sepal width , all species
mean (x [51:100 ,3]) # mean of petal length , versicolor

## ----Convert matrix to a three dimensional array-------------------------
y <- array (x, dim=c(50 , 3, 4))
mean (y[ , ,2]) # mean of sepal width , all species
mean (y[ ,2 ,3]) # mean of petal length , versicolor

y <- array (c(x[1:50 ,] , x [51:100 ,] , x [101:150 ,]) , dim=c(50 , 4, 3))
mean (y[ ,2 ,]) # mean of sepal width , all species
mean (y[ ,3 ,2]) # mean of petal length , versicolor

## ----Lists---------------------------------------------------------------
w <- wilcox.test ( rnorm (10) , rnorm (10 , 2))
w # print the summary
w$statistic #stored in object w
w$p.value

## ----Example 1.6 (A list of names)---------------------------------------
a <- matrix ( runif (8) , 4, 2) #a 4x2 matrix
dimnames (a) <- list (NULL , c("x", "y"))
a
# if we want row names
dimnames (a) <- list ( letters [1:4] , c("x", "y"))
a

# another way to assign row names
row.names (a) <- list ("NE", "NW", "SW", "SE")
a

## ----Reading Data from External Files------------------------------------
 

## ----Graphics,eval=FALSE-------------------------------------------------
## plot.new() #if a plot is not open
## legend(locator(1),as.character(0:25),pch=0:25)#display symbol types
## #then click to locate the legend
## legend(locator(1),as.character(0:25),pch=0:25)#display line types
## legend(locator(1),as.character(0:8),lwd=20,col=0:8)#display color

## ----Other colors and color palettes,eval=FALSE--------------------------
## plot.new()
## palette(rainbow(15))
## legend(locator(1), as.character(1:15), lwd=15, col=1:15)


```

```{r}
########################################
##R Programming Session 1
#########################################

##data entry from keyboard

ages <- c(30, 40, 55, 46, 57)
ages
##input data from spreadsheet

mydata <- read.csv("data/chickwt.csv")
summary(mydata)

##statistical summaries

mean(mydata$weight)

var(mydata$weight)

##modelling. Note the outer parentheses which tell R
##to print the result as well as storing it

(mylm <- lm(weight ~ feed, mydata))

##graphical

plot(density(mydata$weight))

plot(mylm)

boxplot(len ~ dose, data = ToothGrowth,
        boxwex = 0.25, at = 1:3 - 0.2,
        subset = supp == "VC", col = "yellow",
        main = "Guinea Pigs' Tooth Growth",
        xlab = "Vitamin C dose mg",
        ylab = "tooth length",
        xlim = c(0.5, 3.5), ylim = c(0, 35), yaxs = "i")
boxplot(len ~ dose, data = ToothGrowth, add = TRUE,
        boxwex = 0.25, at = 1:3 + 0.2,
        subset = supp == "OJ", col = "orange")
legend(2, 9, c("Ascorbic acid", "Orange juice"),
       fill = c("yellow", "orange"))

## new functions

mycolMeans <- colMeans
fix(mycolMeans)
#############
mymat <- data.matrix(mydata)
mymat[70, 1] <- NA
mycolMeans(mymat)
colMeans(mymat)
colMeans(mymat, na.rm=TRUE)
##help

 
## A brief demo

2 + 3

x <- 2 + 3

exp(-4 * 4 / 2) / sqrt(2 * pi)

dnorm(4, 0, 1) ## or dnorm(0, 4, 1)

##introductory session

library("MASS")
 

x <- rnorm(1000)
y <- rnorm(1000)
truehist(c(x, y + 3), nbins=25)
##save objects
y <- 3
save(x, y, file = "xy.RData")
save.image()
load("xy.RData")
##packages

library("car")

```
  

```{r}
##############################################################
##R Programming Session 2
###############################################################

x <- seq(0, 0.5, 0.1) ##generate a sequence from 0 to 0.5 in steps of 0.1
x ##Look at x
##is x equal to (0, 0.1, 0.2, 0.3, 0.4, 0.5)?
##To find out, type
x == c(0, 0.1, 0.2, 0.3, 0.4, 0.5)

##

myvar <- function(x) (sum(x^2) - length(x) * mean(x)^2) / (length(x) - 1)
x <- seq(1:100)
myvar(x)
var(x)
x <- seq(1:100) + 10000000000
myvar(x)
var(x)
##
##
### Inf, NaN

as.numeric(c("a", "1"))
x / 0
log(-x)
#######

age <- runif(100) * 50
table(cut(age, c(0, 10, 20, 30, 40, 50)))

##why is this different from that on the slide?
##data.frame
## copy the file from the webpage to the folder where you are running R

head(mydata <- read.csv("data/chickwt.csv"))

##list
################
Empl <- list(employee = "Anna", spouse = "Fred",
children = 3, child.ages = c(4, 7, 9))

Empl
Empl[[1]]
Empl$spouse

Empl[[4]]
#Empl[[2:4]] 
## note the error message in previous
length(Empl[[4]])
ls()
str(Empl)

########################
##operators
#############
3^2
10 %% 3
10 %/% 3

a <- matrix(1:4, nrow=2)
b <- matrix(c(2, 1, 2, 4), nrow=2)
a %*% b
###########################
x <- c(1, 2, 3, 4)
y <- c(5, 6)
x + 3
#(4,5,6,7) and
x + y
#(6,8,8,10)

#### logical operators

x <- c(TRUE, FALSE, TRUE)
y <- c(FALSE, TRUE, TRUE)

x | y
x || y

x & y
x && y

################

myDate <- as.Date('10-Jan-1993', format="%d-%b-%Y")

class(myDate)
as.numeric(myDate)

##############

## making vectors
c(1, "a")
seq(1, 10, by=2)# gives (1, 3, 5, 7, 9)
rep(c(1, 2), times=3) #gives (1, 2, 1, 2, 1, 2)
rep(c(1, 2), each=3) #gives (1, 1, 1, 2, 2, 2)
rep(c(1, 2), c(2, 3)) #gives (1, 1, 2, 2, 2)
##

paste(c(1, 2), c('x', 'y', 'z'))
paste(c(1, 2), c('x', 'y', 'z'), collapse=' ')

sort(mydata$weight)
sort(mydata$weight, decreasing=TRUE)

table(rpois(20, 5))

```


```{r}
#########################################################
##R Programming Session 3
##########################################################
mydata <- read.csv('data/chickwt.csv')
summary(mydata)
mydata <- read.csv("data/chickwt.csv",
				   as.is=TRUE)
summary(mydata)
mydata[, 2] <- factor(mydata[, 2])
summary(mydata)

mydata <- read.csv("data/chickwt2.csv")
summary(mydata)
mydata[, 2] <- factor(mydata[, 2])
summary(mydata)

(mymat <- matrix(1:12, 3, 4))
##Entries go down columns unless you
##specify byrow=TRUE.
mymat <- matrix(1:12, 3, 4, byrow=TRUE)
mymat
dim(mymat) #

myarr <- mymat
dim(myarr) <- c(3, 2, 2)
myarr[, , 1]

### indexing vectore
x <- c(2, 4, 6, 8, 10, 12)
names(x) <-
	c("E1", "E2","E3", "E4", "E5", "E6")
x[c(1, 3, 6, 5)]  #2, 6, 12, 10
use <- c(rep(TRUE, 3), FALSE, TRUE, FALSE)
x[use] ##will give 2, 4, 6, 10
x[-c(1:3)] #will give 8, 10, 12
x[c("E1", "E3")]
x[] <- 0

##recycling, extending, extracting outside
x[c(1, 3)] <- 4.5
x[10] <- 8
x[11]
x
### indexing matrices and arrays

mymat[1:2, -2]
mymat[1:2, 1] #is a vector

mymat[1:2, 1, drop=FALSE] # is a matrix

mymat[mymat > 1] <- NA
mymat <- matrix(1:12, 3, 4)

(mysub <- cbind(rep(1, 3), 2:4))
mymat[mysub] <- 2:4
mymat
mymat <- matrix(1:12, 3, 4)

mydata <- read.csv('data/chickwt.csv')

mydata[1:2, 1] #is a vector
mydata[1, 1:2] #is a data.frame
class(mydata[1:2, 1])
class(mydata[1, 1:2])
mydata[mydata$weight > 400,]

##functions
x <- rnorm(100, mean=0.3, sd=1.2)
std.dev <- function(x) sqrt(var(x))
t.test.p <- function(x, mu=0)
{
    n <- length(x)
    t <- sqrt(n) * (mean(x) - mu) /
		std.dev(x)
    2 * (1 - pt(abs(t), n - 1))
}

std.dev(x)
t.test.p(x)
t.test.p(mu=1, x=x)
t.test.p(x, 1)

myfn <- function(n=100)
{
    tmp <- rep(NA, 3)
    tmp[1] <- mean(runif(n))
    tmp[2] <- mean(runif(n))
    tmp[3] <- mean(runif(n))
    mean(tmp[tmp > .2])
}
set.seed(1)
myfn()
myfn(1000)

myfna <- function(n=100)
{
    tmp <- rep(NA, 3)
    x <- mean(runif(n))
    if (x > 0.2)
        tmp[1] <- x
    x <- mean(runif(n))
    if (x > 0.2)
        tmp[2] <- x
    x <- mean(runif(n))
    if (x > 0.2)
        tmp[3] <- x
    mean(tmp, na.rm=TRUE)
}
set.seed(1)
myfna()
myfna(1000)
myfn1 <- function(obs=10, n=100)
{
    x <- rep(NA, n)
    for (i in 1:n)
    {
        tmp <- runif(obs)
        x[i] <- mean(tmp)
    }
    c(mn=mean(x), std=sd(x))
}
set.seed(1)
myfn1()
myfn1(1000)

myfn2 <- function(obs=10)
{
    x <- runif(obs)
    while(mean(x) < 0.45)
    {
        obs <- 2 * obs
        x <- runif(obs)
    }
    list(mn=mean(x), std=sd(x), obs=obs)
}

myfn3 <- function(obs=10)
{
    repeat
    {
        x <- runif(obs)
        if (mean(x) >= 0.45)
            break
        obs <- 2*obs
    }
    list(mn=mean(x), std=sd(x), obs=obs)
}

set.seed(1)
myfn2()
set.seed(1)
myfn3()
myfn3(20)

####
x <- c(0, 1, 1, 2)
y <- c(44, 45, 56, 77)

z <- rep(NA, 4)
for (i in 1:length(x))
{
    if (x[i] > 0)
        z[i] <- y[i] / x[i]
    else
        z[i] <- y[i] / 99
}
z
#or
(z <- ifelse(x > 0, y / x, y / 99))
##
#######
mylist <- list(x=x, y=y, mymat=mymat)
lapply(mylist, length)
## note length of a matrix
sapply(mylist, length)

apply(mymat, 2, sum, na.rm=TRUE)
###tapply
tapply(mydata$weight, mydata$feed, mean)
by(mydata[, 'weight'], mydata$feed, mean)
(mylist <- split(mydata, mydata$feed))

par(mfrow=c(2,3))
tmp <- mapply(function(x, y) hist(x$weight,
						   main = y),
	   mylist, names(mylist))
##

cbind(c(1, 2, 3), c(4, 5, 6))
rbind(c(1, 2, 3), c(4, 5, 6))
df1 <- mylist[[1]][, c(2, 1)]
df2 <- mylist[[2]]
rbind(df1, df2)
cbind(df1[1:10, ], df2)

###########
##merge#


authors <- data.frame(
    surname = I(c("Tukey", "Venables",
	"Tierney", "Ripley", "McNeil")),
    nationality = c("US", "Australia",
	"US", "UK", "Australia"),
    deceased = c("yes", rep("no", 4)))
books <- data.frame(
    name = I(c("Tukey", "Venables",
	"Tierney", "Ripley", "Ripley",
	"McNeil", "R Core")),
    title = c("Exploratory Data Analysis",
              "Modern Applied Statistics ...",
	"LISP-STAT", "Spatial Statistics",
	"Stochastic Simulation",
              "Interactive Data Analysis",
              "An Introduction to R"),
    other.author = c(NA, "Ripley", NA, NA,
	NA, NA, "Venables & Smith"))

(m1 <- merge(authors, books,
			 by.x = "surname",
			 by.y = "name"))
(m2 <- merge(books, authors,
			 by.x = "name",
			 by.y = "surname"))
## "R core" is missing from authors and
## appears only here :
merge(authors, books, by.x = "surname",
	  by.y = "name", all = TRUE)
##matrices
mymat <- matrix(1:12, nrow=3, )
mymat
mymat2 <- matrix(1:12, nrow=3, byrow=TRUE)
mymat2
mymat + mymat2
mymat %*% t(mymat2)

mysq <- matrix(rnorm(9), nrow=3)
solve(mysq)

mysym <- mysq
mysym[lower.tri(mysym)] <-
	mysym[upper.tri(mysym)]
eigen(mysym)
colSums(mymat)

```

### Graphics

```{r}
###################################################################
## R Programming Session 4
##################################################################
x <- 1:10
y <- x^2
plot(y)
plot(list(x=x, y=y))
plot(cbind(x, y))
plot(sin, -pi, pi)
xy <- data.frame(a=x, b=y)
plot(b ~ a, data=xy)
plot(b ~ a, data=xy, type='p')
plot(b ~ a, data=xy, type='l')
plot(b ~ a, data=xy, type='b')
plot(b ~ a, data=xy, type='o')
plot(b ~ a, data=xy, type='h')
plot(b ~ a, data=xy, type='s')

#########################
###plot demo 4.1
#########################
par(mfrow=c(2, 2)) ## sets up screen for four plots

plot(cars)
lines(lowess(cars)) ###add lines

plot(sin, -pi, 2*pi)

## Discrete Distribution Plot:
plot(table(rpois(100, 5)), type = "h", col = "red", lwd=10, lend=1,
     main="rpois(100, lambda=5)")## colour, line width, type, title

## Simple quantiles/ECDF, see ecdf() {library(stats)} for a better one:
plot(x <- sort(rnorm(47)), type = "s", main = "plot(x, type = \"s\")")
points(x, cex = .5, col = "dark red") ##add points, cex

###############################
##legends demo 4.2
###############################
par(mfrow=c(1, 1))

## right-justifying a set of labels: thanks to Uwe Ligges
x <- 1:5; y1 <- 1 / x; y2 <- 2 / x
plot(rep(x, 2), c(y1, y2), type="n", xlab="x", ylab="y")
lines(x, y1); lines(x, y2, lty=2)
temp <- legend("topright", legend = c(" ", " "),
               text.width = strwidth("1,000,000"),
               lty = 1:2, xjust = 1, yjust = 1,
               title = "Line Types")
text(temp$rect$left + temp$rect$w, temp$text$y,
     c("1,000", "1,000,000"), pos=2)
###################
## or use locator() for top left corner (note change of xjust parameter
## to get left corner)
###################
plot(rep(x, 2), c(y1, y2), type="n", xlab="x", ylab="y")
lines(x, y1); lines(x, y2, lty=2)
 


#########################
##other types of plots: demo 4.3
#########################
VADeaths
barplot(VADeaths, beside = TRUE,
        col = c("lightblue", "mistyrose", "lightcyan",
                "lavender", "cornsilk"),
        legend = rownames(VADeaths), ylim = c(0, 100))
#############################
boxplot(len ~ dose, data = ToothGrowth,
        boxwex = 0.25, at = 1:3 - 0.2,
        subset = supp == "VC", col = "yellow",
        main = "Guinea Pigs' Tooth Growth",
        xlab = "Vitamin C dose mg",
        ylab = "tooth length",
        xlim = c(0.5, 3.5), ylim = c(0, 35), yaxs = "i")
boxplot(len ~ dose, data = ToothGrowth, add = TRUE,
        boxwex = 0.25, at = 1:3 + 0.2,
        subset = supp == "OJ", col = "orange")
legend(2, 9, c("Ascorbic acid", "Orange juice"),
       fill = c("yellow", "orange"))
 

#########################
## basic lattice plots demo 4.5
########################
library(lattice)
library(MASS)
xyplot(time ~ dist, data=hills)

bwplot(voice.part ~ height, data=singer, xlab="Height (inches)")

stripplot(voice.part ~ jitter(height), data = singer, aspect = 1,
          jitter = TRUE, xlab = "Height (inches)")

## basic - note type 'b' is different in lattice
xyplot(Time ~ Viscosity, data = stormer, groups = Wt, type = "b")

########################
## panel functions demo 4.6
########################
xyplot(time ~ dist, data=hills,
panel=function(x, y, ...){
         panel.xyplot(x, y, ...)
         panel.lmline(x, y, type='l')
     })

xyplot(time ~ dist, data=hills,
panel=function(x, y, ...){
         panel.xyplot(x, y, ...)
         panel.lmline(x, y, type='l')
          panel.abline(lqs(y ~ x), lty=3)
    })

xyplot(time ~ dist, data=hills,
	   panel=function(x, y, ...){
		   panel.xyplot(x, y, ...)
		   panel.lmline(x, y, type='l')
		   panel.abline(lqs(y ~ x), lty=3)
		 ##  panel.identify(x, y, row.names(hills))
	   })
##note the axes used to appear when the identify() finished.
## 2012: this code does not work at this time, so rework the example
## 2013: still not working
trellis.focus("panel", 1, 1)
panel.identify(hills$dist, hills$time, row.names(hills))

#################################
#####conditioning demo 4.7
##############################
##conditioning with factor
bwplot(Age ~ Days | Sex * Lrn, data = quine)

##conditioning with a shingle
## Fertility is a standardized fertility measure
## Education: percentage beyond primary school of army draftees
## Catholic: percentage Catholic as opposed to Protestant (c.1888)
## Examine relationship between Fertility and Education
Cath <- equal.count(swiss$Catholic)
xyplot(Fertility ~ Education | Cath, data=swiss)

#################################
#####conditioning demo 4.8
##############################

##alter the strip labels
##quine: Days absent from school
## Age group Primary (F0), F1, F2, F3
## Lrn Average or slow
bwplot(Age ~ Days | Sex * Lrn, data = quine, strip=strip.custom(style=2))

## and specify the layout
## yield in bushels per acre for 10 varieties at six sites in each of 2 yrs
barchart(yield ~ variety | site, data = barley,
         groups = year, layout = c(1, 6),
         ylab = "Barley Yield (bushels/acre)",
         scales = list(x = list(abbreviate = TRUE,
                       minlength = 5)))

###################################
### Keys demo 4.9
###################################


## Note Multiple variables in formula for grouped displays. See also 4.11
##auto.key
xyplot(Sepal.Length + Sepal.Width ~ Petal.Length + Petal.Width | Species,
       data = iris, scales = "free", layout = c(2, 2),
       auto.key = list(x = .6, y = .7, corner = c(0, 0)))

## time for cylinder to rotate fixed nbr times in response to wt, known visc
xyplot(Time ~ Viscosity, data = stormer, groups = Wt, type = "b",auto.key=TRUE)

##simplekey
mylabels<-paste(c("Weight:   ", "", ""), unique(stormer$Wt), "gms")
 xyplot(Time ~ Viscosity, data = stormer, groups = Wt, type = "b",
    key = simpleKey(columns = 3, text=mylabels))


##alter the character and customize the key
sps <- trellis.par.get("superpose.symbol")
sps$pch <- 1:7
trellis.par.set("superpose.symbol", sps)
xyplot(Time ~ Viscosity, data = stormer, groups = Wt, type = "b",
 key = list(columns = 3, text = list(paste(c("Weight:   ", "", ""),
        unique(stormer$Wt), "gms")),
        points = Rows(sps, 1:3)
 )
)

##################################
###multiple plots on one screen demo 4.10
###################################

plot1 <- xyplot(dist ~ speed, cars)

plot2 <- xyplot(sin(x)~x, data=data.frame(x=seq(-pi, 2 * pi, length=100)),
                type='l')

## Discrete Distribution Plot:
tmp <- table(rpois(100, 5))
plot3 <- xyplot(as.vector(tmp) ~ as.numeric(names(tmp)), type = "h",
                col = "red", lwd=10,
                main="rpois(100, lambda=5)")## colour, line width, type, title

## Simple quantiles/ECDF, see ecdf() {library(stats)} for a better one:
x <- sort(rnorm(47))
plot4 <- xyplot(x ~ 1:47, type = "s", main = "plot(x, type = \"s\")",
       panel=function(x, y){
       panel.xyplot(x, y, type='s')
   panel.xyplot(x, y, type='p', cex=0.5, col='dark red')})

###layout

print(plot1, newpage=TRUE, position=c(0, .5, .5, 1), more=TRUE)
print(plot2, position=c(.5, .5, 1, 1), more=TRUE)
print(plot3, position=c(0, 0, .5, .5), more=TRUE)
print(plot4, position=c(.5, 0, 1, .5))
##or
print(plot1, newpage=TRUE, split=c(1, 1, 2, 2), more=TRUE)
print(plot2, split=c(2, 1, 2, 2), more=TRUE)
print(plot3, split=c(1, 2, 2, 2), more=TRUE)
print(plot4, split=c(2, 2, 2, 2))

############################################
### Extended formula interface: demo 4.11
###########################################
## multiple plots against a factor
stripplot(type ~ RI + Na + Mg + Al + Si + K + Ca + Ba + Fe,
outer=TRUE, data=fgl, scales=list(x="free"), xlab="", cex=0.5)
stripplot(type ~ RI + Na + Mg + Al + Si + K + Ca + Ba + Fe,
outer=FALSE, data=fgl, scales=list(x="free"), xlab="", cex=0.5)
## parallel time series
deathdf <- data.frame(mdeaths, fdeaths)
xyplot(fdeaths + mdeaths ~ 1:72, deathdf, type="l")
xyplot(fdeaths + mdeaths ~ 1:72, deathdf, type="l", outer=TRUE)
## easier way than this?
deathdf2 <- data.frame(deaths=c(mdeaths, fdeaths), obs=rep(1:72, 2),
					   gender=rep(c("m", "f"), each=72))
xyplot(deaths ~ obs, deathdf2, type="l", groups=gender)
## or this: remove the ts class, then reshape?
deathdf3 <- data.frame(mdeaths=as.vector(mdeaths), fdeaths=as.vector(fdeaths))
deathdf3 <- reshape(deathdf3, direction="long", varying=list(1:2))
xyplot(mdeaths ~ id, deathdf3, type="l", groups=time)

############################################
### Example code
###########################################
xyplot(dist ~ speed, cars)

bwplot(count ~ spray, InsectSprays,
  xlab='Type of Spray',
  ylab='Number of Insects',
  main='Insect Spray Data')

xyplot(mpg ~ disp, groups=cyl, data=mtcars,
     xlab='Displacement (cu in.)',
     ylab='Miles/(US)gallon',
     main='Car Road Tests Data',
     auto.key=TRUE)

xyplot(mpg ~ disp, groups=cyl, data=mtcars,
       key=simpleKey(text=unique(as.character(mtcars$cyl)), columns=3))
###########################################################
##update examples
Depth <- equal.count(quakes$depth)

(tmp <- xyplot(lat ~ long | Depth, data = quakes))
(tmp <- update(tmp,
       strip = strip.custom(strip.names = TRUE, strip.levels = TRUE),
       aspect = "iso"))

xyplot(time ~ dist, data=hills)
trellis.focus("panel", 1, 1)
panel.identify(hills$dist, hills$time, row.names(hills))

```

#### Character data

```{r}
########################################
##R Programming Session 6
#########################################

##lengths of strings

length("")

vv <- c("abcd", "def", "k")
##create an empty character vector
vv0 <- vv[vv == "g"]

##length versus nchar

length(vv)

length(vv[1])

length((vv0))

nchar(vv[1])

nchar(vv)

nchar(vv0)

##special characters

a <- 'The aunt\'s pen'
b <- "The aunt's pen"
a
b
a == b

vspecial <- "A \t tab"
print(vspecial)
cat(vspecial)
nchar(vspecial)

##paste
paste(c("X", "Y"), 1:4, sep="")
paste(c("X", "Y"), 1:4, sep="", collapse=" + ")

##substr, substring

substr("abcdef", 2, 4)
substring("abcdef", 2, 4)

substring("abcdef", 1:6, 1:6)
substr("abcdef", 1:6, 1:6) # vector arguments longer than length of string
# are ignored

substr(rep("abcdef", 4), 1:4, 4:5)

rep("abcdef", 4)
cbind(1:4, 4:5)

x <- c("asfef", "yuiop[", "b", "stuff.blah.yech")
substr(x, 2, 5) ## note results where index is not present

substring(x, 2, 4:6) ## uses recycling

substring(x, 2) <- c("..", "+++")
x
##note default final argument to substring is 1000000
##only portion of replacement string is replaced.


##example of strsplit
x <- c(as="asfef", qu="qwerty", "yuiop[", "b",
    "stuff.blah.yech")
strsplit(x, "e")  # split x on the letter e

##example of match
match(1:10, c(1, 3, 5, 9))

1:10 %in% c(1, 3, 5, 9)
pmatch("", "")                # NA
pmatch("m", c("mean", "median", "mode")) # NA
pmatch("med", c("mean", "median", "mode"))# 2
charmatch("", "")                     #1
charmatch("m", c("mean", "median", "mode"))#0
charmatch("med", c("mean", "median", "mode"))#2
charmatch("x", c("mean", "median", "mode"))# NA

##duplicates
pmatch(c("", "ab", "ab"), c("abc", "ab"), dup=FALSE)
pmatch(c("", "ab", "ab"), c("abc", "ab"), dup=TRUE)
##Compare with
charmatch(c("", "ab", "ab"), c("abc", "ab"))

##ambiguous targets
charmatch(c("", "ab", "ab"), c("abc", "ab", "ab"))
pmatch(c("", "ab", "ab"), c("abc", "ab", "ab"), dup=FALSE)
pmatch(c("", "ab", "ab"), c("abc", "ab", "ab"), dup=TRUE)

## print versus cat
x <- c("abc", "dsef\n", "g")
print(x)
cat(x)

## more examples of cat, and other functions
d <- date()
cat("Today's date is:", substring(d, 1, 10),
    substring(d, 21, 24), "\n")
 cat(1, 2, 3, 4, 5, 6, fill = 7, labels = letters)
##format command
format(pi, width=12, nsmall=6)
format(c(pi, pi*pi, pi^3))
format(signif(c(pi, pi * pi, pi^3), digits=3))
format(signif(c(pi, pi * pi, pi^3)), digits=3)

##sprintf also a possibility
##modifying strings
x <- "My string 123"
chartr("yis", "abC", x)
chartr("r-tM", "A-Cw", x)
tolower(x)
toupper(x)
##regular expressions
x <- "My pattern string 32144"
##to demonstrate the matches
gregexpr("n.", x) #. matches any character
gregexpr("[1-3]", x) #[1-3] matches any 1, 2  or 3
gregexpr("n\\>", x) # only consider matches at end of a word
gregexpr("\\<p", x) # only consider matches at end of a word
gregexpr("t{2}", x)  #only matches occurrences of 2 t's
##to demonstrate the behaviour of the different functions
grep("pa", c(x, tolower(x))) ## match in each element
grep("pa", c(x, toupper(x))) ## match in first element
grep("pa", c(x, toupper(x)), value=TRUE) # return values of element that matches
grepl("pa", c(x, toupper(x))) ## match in first element

regexpr("pa", c(x, toupper(x))) #match in first
regexpr("n", c(x, toupper(x))) #match in first
gregexpr("n", c(x, toupper(x))) ##  two matches in first string
##replacement in strings
 x <- "My pattern string 12344"

sub("PA", "pa", c(x, toupper(x)))

sub("n.", "ZZ", c(x, toupper(x))) ## only changes first match

gsub("n.", "XX", c(x, toupper(x))) ## changes both occurrences

###botanical database

###data
mycomments <- c("Salpichroa sp?, Cotopaxi NP, Ecuador",
             "Monnina sp, Cotopaxi NP, Ecuador",
             "Baccharis latifolia, Cotopaxi NP, Ecuador",
               "Calceolaria sp, Cotopaxi NP, Ecuador")
Genus1 <- data.frame(genus=c(" ","Baccharis","Calceolaria"),
                     family=c("Not_Known","Asteraceae","Calceolariaceae"),
                    stringsAsFactors=FALSE)

###extract the botanical name
botname <- ifelse(regexpr(",", mycomments) > -1,
                 substring(mycomments, 1, regexpr(",", mycomments) - 1),
                 mycomments)
###or
mybotnamelist <- strsplit(mycomments, ",")
mybotnamelist[c(1, 2)]
sapply(mybotnamelist[c(1, 2)], "[", 1)
botname <- sapply(mybotnamelist,"[", 1)
botname
###extract the genus (first part of name)
genus <- substring(botname, 1, regexpr(" ", botname) - 1)
genus
###create a data.frame of genera
genusdf <-  data.frame(genus=unique(genus), stringsAsFactors=FALSE)
###merge with the previous version, Genus (has family field too)
Genusall <- merge(Genus1, genusdf, all=TRUE)
Genusall
###outside R, fill in the families for the new genera, then read in again
### new data frame Genus
Genus2 <- data.frame(genus=c(" ", "Baccharis", "Calceolaria", "Monnina",
                    "Salpichroa"),
                     family=c("Not_Known", "Asteraceae", "Calceolariaceae",
                     "Polygalaceae", "Solanaceae"),
                   stringsAsFactors=FALSE )
###find the family for the new photos
Family <- Genus2[,2][match(genus, Genus2[,1], nomatch=1)]
cbind(genus, Family)

```


Attribute
http://portal.stats.ox.ac.uk/userdata/ruth/APTS2013/  
http://computationalfinance.lsi.upc.edu/?page_id=28 
```{r}
############################
## R programming session 8
############################

##attributes

library(MASS)
lh ### a time series object

attributes(lh)

attr(lh, "tsp")

attr(lh, "ownership") <- "mine"

attributes(lh)


## generic functions

methods(summary)

methods(class="lm")

mylm <- lm(weight ~ feed, data=chickwts)

summary(mylm)

summary(chickwts)

##summary.aov, summary.lm:

op <- options(contrasts=c("contr.helmert", "contr.poly"))

npk.aov <- aov(yield ~ block + N * P * K, npk)
summary(npk.aov)
summary.lm(npk.aov)

options(op)# restore to previous

## s4 slots
library(Matrix)
(I3 <- Matrix(diag(3)))# identity, i.e., unit "diagonalMatrix"
class(I3)
str(I3)
getSlots("ddiMatrix")
slotNames(I3)
I3@Dim

myFn <- function(x, ...) UseMethod("myFn")
myFn.default <- function(x, ...)
              print("in myFn.default")
myFn("test")

myFn.character <- function(x, ...)
{
    print(x)
    NextMethod(x)
    print(c(x, 2))
}
myFn("test")

## search paths, conflicts
search()
conflicts()
conflicts(detail=TRUE)
## finding things

find("lh")
exists("lh")

## local copies

lm <- lm
conflicts()

##evaluation environments

f1 <- function()
{
    print(x)
}
f2 <- function()
{
   x <- 2
   f1()
}

f2()  ## gives error - exactly what depends on whether Matrix is loaded.

## functions with environments
s <- 5; x <- 6
f <- function()
{
    ## define function g: prints s and x,
    ## returns NULL
    ## will get environment from f
    g <- function()
    {
        cat("g: s=", s, "x=", x, "\n")
    }
    ## define function j which:
    ## 1) defines function k,
    ## 2) creates s with value 1,
    ## 3) returns result of running k
    j <- function()
    {
        ## define function k which returns a
        ## function with an environment
        ## containing s with value 1
        k <- function()
        {
            function() cat("anon in k: s=",
                           s, "x=", x, "\n")
        }
        s <- 1
        k()
    }
    ## create s and x in the environment of f
    ## (and g and j)
    s <- 3
    x <- 4
    cat("run j:\n")
    print(j())
    cat("\nrun return val of j:\n")
    j()()
    cat("result of running g:\n")
    g()
}

f()
## want to see the boxcox function (need MASS here)
boxcox
## not very helpful
methods(boxcox)
## non-visible
getAnywhere(boxcox.lm)
MASS:::boxcox.lm

##returning values from functions

a <- NULL
a$start <- 1
myfn <- function(x)
{
    x$start <- 2
    return(x)
}
(a <- myfn(a))

```

https://rpubs.com/ktmud/ggplot2  


    manufacturer - car manufacturer
    model - car model
    displ - engine displacement, in litres
    year - year of manufacture
    cyl - number of cylinders
    trans - type of transmission
    drv - f = front-wheel drive, r = rear wheel drive, 4 = 4wd
    cty - city miles per gallon
    hwy - highway miles per gallon
    fl - fuel type
    class - “type” of car



```{r}
library(ggplot2)
p <- ggplot(data=mpg, aes(displ, hwy))
p
```

```{r}
p <- ggplot(data=mpg, aes(x=displ, y=hwy))
p + geom_point()
```



Essential Grammatical Elements

    data - data being plotted (R data frame)
    aesthetics - visual characteristics that represent data
    geoms - geometric objects representing the data

All Grammatical Elements

    data - data being plotted (R data frame)
    aesthetics - visual characteristics that represent data
    geoms - geometric objects representing the
    scales - how visual characteristics are converted to display values
    stats - statistical transformations summarizing data
    coordinate system - space used for projecting data
    facets - describes data splits in subsets displayed as multiple graphs
    themes - all non-data ink


Add color aesthetics (type of a class)


```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(color=class))
```

Add color aesthetics (number of cylinders) - continuous  

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(color=cyl))
```

Add color aesthetics (number of cylinders) - discrete using a factor  

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(color=factor(cyl)))
```

Add size aesthetics (number of cylinders)  

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(size=cyl))
```

Add size aesthetics 

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(color=class, size=cyl))
```

Add shape aesthetics 

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(aes(shape=factor(cyl)), size=4)
```

Add more shape aesthetics 

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(colour="black", size=4.5) +
  geom_point(colour="white", size=4) +
  geom_point(aes(shape = factor(cyl)))
```

Show the pimped shapes in legend  

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point(colour="black", size=4.5, show.legend=T) +
  geom_point(colour="white", size=4, show.legend=T) +
  geom_point(aes(shape = factor(cyl)))
```

https://github.com/clayford?tab=repositories  
List of aesthetics

    x
    y
    color
    fill
    size
    alpha
    linetype
    labels

Geoms

```{r}
p <- ggplot(mpg, aes(x=displ, y=hwy))
p + geom_point() + geom_line()
```
https://rpubs.com/ktmud/ggplot2 
https://github.com/anhnguyendepocen/IntroEDA_Rggplot2  
https://rpubs.com/JanpuHou/326048 
https://web.stanford.edu/class/stats366/Gibbs.html 
http://rstudio-pubs-static.s3.amazonaws.com/279858_010f9da7c8d744988019397e3fe51cb2.html  
http://www.est.uc3m.es/icascos/eng/simulation_notes/mcmc-techniques.html  
https://fabiandablander.com/r/Spike-and-Slab.html 
